Comparing Facebook's Prophet Forecasting to an ARIMA Model: A Diebold-Mariano Evaluation of the Top40 Index

### 2017-10-14 template/Draft_Paper.Rmd
Tables in markdown, some, but not all plots in markdown, fixed issue with dates on figure 1, in-sample data with dplyr

The data set used throughout this paper comprises of the daily closing prices of the equally-weighted FTSE/JSE Top40 price index for the five year period from 31/07/2012 to 30/07/2017. This equates to a total of 1766 trading days excluding weekends and holidays. This period was chosen for its recency and to avoid exposing the results to the noise associated with the high volatility experienced in the stock market during the 2008 financial crisis. The aim is to compare the predictive accuracy of the two models under standard market conditions before other confounding factors can be closely analysed.  Figure 3 shows the actual realisations of the price index data which was sourced from Datastream.

![Figure 2](plot of Price index.png){ width=100% }  
*Figure 3: Realisation of the FTSE/JSE Top40 Price Index over the period of study*

@@

In this paper, the **log return**(which will be used interchangeably with **return**), over the period $[t,t+1]$ is denoted by $r_t$, and was calculated by evaluating;  
$$
r_t = log(\frac{P_{t+1}}{P_t})  \label{eq1}
$$  
Where $P_t$ and $P_{t+1}$ represent the price of the index at time $t$ and $t+1$ respectively.

@@

@@

Observations from the above plots informed the choice of candidate *ARIMA(p,d,q)* models to fit to the data. The candidate models that were fit to the in-sample data are shown in Table 1 with their respective AIC values. The candidate model with the lowest AIC is the $ARIMA(2,0,2)$. This model was thus selected as the ARIMA forecasting model.  
![](AIC.png){ width=100% }  
*Table 1: AIC for Candidate ARIMA models*

## Forecast Error Evaluation
The results of model estimation and forecast evaluation using the rolling window for h-step forecasts yielded forecasts errors with roughly equal distributions. This is visible from the boxplot in Figure 5. The presence of outliers seems consistent across the different models **and** forecasting horizons. Furthermore, variations in accuracy across different forecasting horizons *for each* model is not observable in Figure 5. The mean error and spread of both models across different time horizons appears to be constant. 
![](Forecast error boxplot.png){ width=105% }  
*Figure 5: Boxplot illustrating distribution of rolling forecast errors for different time horizons, h*

@@

![ErrorStats](error stats.png){ width=100% }  
*Table  2 : Forecast error diagnostic statistics over different forecasting horizons*

>>>>

The data set used throughout this paper comprises of the daily closing prices of the FTSE/JSE Top40 price index for the five year period from 31/07/2012 to 31/07/2017. This equates to a total of 1305 trading days excluding weekends and holidays. This period was chosen for its recency and to avoid exposing the results to the noise associated with the high volatility experienced in the stock market during the 2008 financial crisis. The aim is to compare the predictive accuracy of the two models under standard market conditions before other confounding factors can be closely analysed.  Figure 3 shows the actual realisations of the price index data which was sourced from Datastream.

```{r, echo=FALSE}
#Import, transform and plot data
library(dplyr)
library(readxl)
top40index <- read_excel("~/Honours_project/Draft-Paper/top40index.xlsx", 
                         col_types = c("date", "numeric"))
index <- top40index%>%distinct(date, .keep_all = TRUE) #Removing duplicate dates

plot(index, type = "l",
     main = "FTSE/JSE Top40 Index 07/2012 - 07/2017")
```
*Figure 3: Realisation of the FTSE/JSE Top40 Price Index over the period of study*

@@

In this paper, the **log return** (which will be used interchangeably with **return**), over the period $[t,t+1]$ is denoted by $r_t$, and was calculated by evaluating;  
$$
r_t = log(\frac{P_{t+1}}{P_t})  \label{eq1}
$$  
Where $P_t$ and $P_{t+1}$ represent the price of the index at time $t$ and $t+1$ respectively.

@@

```{r, eval=FALSE}
# Outliers
boxplot(c(index[,2]), names = c("price index"), main = "Boxplot of Price Index")  #Comment : looks like there are some outliers eventhough outliers are 'relatively' clustered
log.returns <- CalculateReturns(as.ts(top40index[,2]), method = "log")[-1] 
boxplot(log.returns, names = c("log returns"), main = "Boxplot of Log Returns on the Price Index")

#Formal tests for Outliers in Time series --------------------------

library(tsoutliers)  # conducts multiple tests for outliers #Reference Chen and Liu;s as authors : https://stats.stackexchange.com/questions/104882/detecting-outliers-in-time-series-ls-ao-tc-using-tsoutliers-package-in-r-how
library(ggpubr)
log.returns.outlier.detection <- tsoutliers::tso(as.ts(log.returns),types = c("AO","IO","LS","TC","SLS"), maxit.iloop = 10)
log.return.outlier.detection  #Several outliers detected
#Further investigation shows that these outliers occur when the return is large in abs value, because this is due to natural market activity and the outliers do not reflect in the boxplot of the price index, the outliers have been kept in the data. See : log.returns[c(299,302,348,1144,1156,1540)]

@@

Observations from the above plots informed the choice of candidate *ARIMA(p,d,q)* models to fit to the data. The candidate models that were fit to the in-sample data are shown in Table 1 with their respective AIC values. The candidate model with the lowest AIC is the $ARIMA(2,0,2)$. This model was thus selected as the ARIMA forecasting model.

```{r ShortTable, results = 'asis'}

library(xtable)

candidate <- c("(0,0,0)","(2,0,0)","(0,0,2)","(2,0,2)")
aic.values <- c(-8804,-8808,-8808,-8817)

aic.mat <- matrix(data = c(candidate,aic.values),ncol = 2)
dimnames(aic.mat) <- list(c(" "," "," "," "),   #row names
                          c("ARIMA(p,d,q) model","AIC"))  #column names
data = aic.mat %>% tbl_df()

table <- xtable(data, caption = "AIC for Candidate ARIMA models")
  print.xtable(table, 
             # tabular.environment = "longtable",
             floating = TRUE,
             table.placement = 'H', 
             # scalebox = 0.3, 
             comment = FALSE,
             include.rownames = getOption("xtable.include.rownames", FALSE), #exclude row number
             caption.placement = 'bottom'
             )

```
## Forecast Error Evaluation
The results of model estimation and forecast evaluation using the rolling window for h-step forecasts yielded forecasts errors with roughly equal distributions. This is visible from the boxplot in Figure 5. The presence of outliers seems consistent across the different models **and** forecasting horizons. Furthermore, variations in accuracy across different forecasting horizons *for each* model is not observable in Figure 5. The mean error and spread of both models across different time horizons appears to be constant. 
![](Forecast error boxplot.png){ width=105% }  
*Figure 5: Boxplot illustrating distribution of rolling forecast errors for different time horizons, h*

@@

```{r ForecastErrorStatisticsTable, results = 'asis'}

library(xtable)

c1 <- c("1"," ","5"," ","20"," ")
c2 <- rep(c("ARIMA","Prophet"),3)
c3 <- c('-0.0004110','0.0000176','-0.0003635','0.004445','-0.0001769','-0.0000406')
c4 <- c(0.0103,0.0104,0.0103,0.0103,0.0102,0.0102)
c5 <- c(0.007718,0.007749,0.007692,0.007709,0.007649,"0.007660")

error.stats.mat <- matrix(data = c(c1,c2,c3,c4,c5),ncol = 5)
dimnames(error.stats.mat) <- list(c(" "," "," "," "," "," "),   #row names
                          c("Forecast Horizon, h","Forecast Model"," ME ","RMSE","MAE"))  #column names
err.stat.data = error.stats.mat %>% tbl_df()

error.table <- xtable(err.stat.data, caption = "Forecast error diagnostic statistics over different forecasting horizons")
  print.xtable(error.table, 
             # tabular.environment = "longtable",
             floating = TRUE,
             table.placement = 'H', 
             # scalebox = 0.3, 
             comment = FALSE,
             include.rownames = getOption("xtable.include.rownames", FALSE), #exclude row number
             caption.placement = 'bottom'
             )
